Composable data visualizations

### 2020-05-09 wip-compost/paper.tex
New implementation approach, abstractions

Our design focuses on two-dimensional charts with X and Y axes. Values mapped to those axes
can be either categorical (such as different political parties or countries) or continuous
(such as number of votes or exchange rates). The mapping from categorical and continuous values
to exact positions on the chart is done automatically. For continuous values, this simply means
applying a linear transformation. For categorical values, the mapping is more difficult.

@@

Figure~\ref{fig:scales} illustrates the two kinds of values using the axes from the UK
election results chart. In Figure~\ref{fig:shape}, we define a value $v$ as either a continuous value
$\kvd{cont}~n$ containing any number $n$ or a categorical value $\kvd{cat}~c, r$, consisting
of a categorical value $c$ (implemented as a string) and a ratio $r$ between $0$ and $1$.
%
\begin{figure}
\begin{equation*}
\begin{array}{rcl}
v & = & \kvd{cat}~c, r \\
  & | & \kvd{cont}~n\\
  ~
\end{array}
\qquad
\begin{array}{rclcl}
s & = &\kvd{line}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] & | & \kvd{overlay}~[\,s_1, \ldots, s_n\,]\\
 & | & \kvd{fill}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] & | &\kvd{axis}_{l/r/t/b}~s\\
 & | & \kvd{text}~\gamma, v_x, v_y, t & | &\kvd{padding}~n_t,n_r,n_b,n_l,s
\end{array}
\end{equation*}
\caption{Core primitives of the Compost domain-specific language. Values $v$ are either categorical
  or continuous; a shape $s$ is then defined as a simple recursive algebraic data type.}
\label{fig:shape}
\end{figure}

@@

Now that we know how Compost represents values, we can define the basic elements of its
domain-specific langauge. A chart is represeted by the shape $s$ defined in Figure~\ref{fig:shape}.
A primitive shape can be a text label, a line connecting a list of points or a filled polygon
defined by a list of points. The position of points is specified by X and Y coordinates,
which can be either categorical or continuous values. For text, line and polygon, we also include
a parameter $\gamma$ that specifies the element color.

@@

\noindent
Aside from iterating over all available parties and splitting the bar, the example also adds padding
around the bars. A padding is specified in pixels rather than in terms of domain values. This is
sometimes preferrable over, for example, drawing a bar using a range from $\num{0.05}$ to $\num{0.5}$.
One remaining feature of the chart in Figure~\ref{fig:charts} that is still missing is the caption.
We will add this in Section~\ref{sec:abstractions}.

@@

For completeness, the following code snippet shows how to construct the full currency exchange
rate chart shown in Figure~\ref{fig:lines} (right), including the blue and red background:
%
\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{brexitRate}~(\ident{lo}, \ident{hi})~\ident{rates}~=\kvd{overlay}~[\\
\quad \kvd{fill}~\strf{\#1F77B460}, [\;\,\kvd{cont}~\num{0}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{hi}, \kvd{cont}~\num{0}, \kvd{cont}~\ident{hi}\;],\\
\quad \kvd{fill}~\strf{\#D6272860}, [\;\,\kvd{cont}~\num{16}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{44}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{44}, \kvd{cont}~\ident{hi}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{hi}\;],\\
\quad \kvd{line}~\strf{\#202020}~\ident{rates}~]\\[0.5em]
\kvd{overlay}~[~
\kvd{nest}_y~(\kvd{cont}~\num{0}),(\kvd{cont}~\num{50}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\ident{brexitRate}~(\num{1.25}, \num{1.50})~\ident{gbpusd}))))\\
\hspace{3.68em} \kvd{nest}_y~(\kvd{cont}~\num{50}),(\kvd{cont}~\num{100}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\ident{brexitRate}~(\num{1.15}, \num{1.30})~\ident{gbpeur}))))~]\\
\end{array}
\end{equation*}

@@

\section{Definining abstractions}
\label{sec:abstractions}

\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{title}~t~s~=\ident{overlay}~[\\
\quad \kvd{nest}_x~(\kvd{cont}~\num{0}), (\kvd{cont}~\num{100}),
  (\kvd{nest}_y~(\kvd{cont}~\num{0}), (\kvd{cont}~\num{15}), \\
\quad\quad \kvd{explicitScale}_x~(\kvd{continuous}~\num{0},\num{100}),~(\kvd{explicitScale}_y~(\kvd{continuous}~\num{0},\num{100}),\\
\quad\qquad \kvd{text}~\strf{\#000000}, (\kvd{cont}~\num{50}), (\kvd{cont}~\num{50}),~t)~)\\
\quad \kvd{nest}_x~(\kvd{cont}~\num{0}), (\kvd{cont}~\num{100}),
  (\kvd{nest}_y~(\kvd{cont}~\num{15}), (\kvd{cont}~\num{100}),~s)~]
\end{array}
\end{equation*}

\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{pairplot}~\ident{attrs}~\ident{data}~=~\kvd{overlay}~[\\
\quad \kvd{for}~x~\kvd{in}~\ident{attrs}~\rightarrow~\kvd{for}~y~\kvd{in}~\ident{attrs}~\rightarrow\\
\qquad \kvd{nest}_x~(\kvd{cat}~x, 0), (\kvd{cat}~x, 1), (\kvd{nest}_y~(\kvd{cat}~y, 0), (\kvd{cat}~y, 1), \\
\quad\qquad (\kvd{axis}_l~(\kvd{axis}_b~(\kvd{overlay}~[~\fkvd{for}~v~\kvd{in}~\ident{data}~\rightarrow~\kvd{bubble}~(\ident{get}~x~v), (\ident{get}~y~v), \num{1}, \num{1} ])))~)]\\
\end{array}
\end{equation*}

let
  Shape.Layered [
    for x in ["sepal.length"; "sepal.width"; "petal.length"; "petal.width" ] do
      for y in ["sepal.length"; "sepal.width"; "petal.length"; "petal.width" ] do
        yield Shape.OuterScale(Some(Categorical [| ca x |]), Some(Categorical [| ca y |]),
          Shape.Axes(false, false, true, true, Shape.Layered [
            for v, k in iris ->
              Derived.StrokeColor(colors.[k], Shape.Bubble(numv v.[x], numv v.[y], 1., 1.))
          ]))
  ] |> render "out3"

>>>>

Our design focuses on two-dimensional charts with X and Y axes. Values mapped to those axes
can be either categorical (e.g.~political parties, countries) or continuous
(e.g.~number of votes, exchange rates). The mapping from categorical and continuous values
to positions on the chart is done automatically. For continuous values, this involves
applying a linear transformation. For categorical values, the mapping is more difficult.

@@

Figure~\ref{fig:scales} illustrates the two kinds of values using the axes from the UK
election results chart. In Figure~\ref{fig:shape}, we define a value $v$ as either a continuous value
$\kvd{cont}~n$ containing any number $n$ or a categorical value $\kvd{cat}~c, r$, consisting
of a categorical value $c$ (implemented as a string) and a ratio $r$ between $0$ and $1$.
%
\begin{figure}
\begin{equation*}
\begin{array}{rcl}
v & = & \kvd{cat}~c, r \\
  & | & \kvd{cont}~n\\
  ~
\end{array}
\qquad
\begin{array}{rclcl}
s & = &\kvd{line}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] & | & \kvd{overlay}~[\,s_1, \ldots, s_n\,]\\
 & | & \kvd{fill}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] & | &\kvd{axis}_{l/r/t/b}~s\\
 & | & \kvd{text}~\gamma, v_x, v_y, t & | &\kvd{padding}~n_t,n_r,n_b,n_l,s\\
 & | & \kvd{bubble}~\gamma, v_x, v_y, w, h
\end{array}
\end{equation*}
\caption{Core primitives of the Compost domain-specific language. Values $v$ are either categorical
  or continuous; a shape $s$ is then defined as a simple recursive algebraic data type.}
\label{fig:shape}
\end{figure}

@@

Now that we know how Compost represents values, we can define the basic elements of its
domain-specific langauge. A chart is represeted by the shape $s$ defined in Figure~\ref{fig:shape}.
A primitive shape can be a text label, a line connecting a list of points, a filled polygon
defined by a list of points or a bubble at a given point with a given width and height.
The position of points is specified by X and Y coordinates, which can be either categorical or
continuous values. For text, line, polygon and bubble, we also include a parameter $\gamma$
that specifies the element color. The width and height of a bubble is given in pixels.

@@

\noindent
Aside from iterating over all available parties and splitting the bar, the example also adds padding
around the bars. A padding is specified in pixels rather than in terms of domain values. This is
sometimes preferrable over, for example, drawing a bar using a range from $\num{0.05}$ to $\num{0.5}$.
The chart is still missing a title, which we add in Section~\ref{sec:abstractions}.

@@

For completeness, the following code snippet shows how to construct the full currency exchange
rate chart shown in Figure~\ref{fig:lines} (right), including the blue and red background:
%
\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{xrate}~(\ident{lo}, \ident{hi})~\ident{rates}~=\kvd{overlay}~[\\
\quad \kvd{fill}~\strf{\#1F77B460}, [\;\,\kvd{cont}~\num{0}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{hi}, \kvd{cont}~\num{0}, \kvd{cont}~\ident{hi}\;],\\
\quad \kvd{fill}~\strf{\#D6272860}, [\;\,\kvd{cont}~\num{16}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{44}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{44}, \kvd{cont}~\ident{hi}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{hi}\;],\\
\quad \kvd{line}~\strf{\#202020}~\ident{rates}~]\\[0.5em]
\kvd{overlay}~[~
\kvd{nest}_y~(\kvd{cont}~\num{0}),(\kvd{cont}~\num{50}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\ident{xrate}~(\num{1.25}, \num{1.50})~\ident{gbpusd}))))\\
\hspace{3.68em} \kvd{nest}_y~(\kvd{cont}~\num{50}),(\kvd{cont}~\num{100}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\ident{xrate}~(\num{1.15}, \num{1.30})~\ident{gbpeur}))))~]\\
\end{array}
\end{equation*}

@@

\begin{figure}
  \includegraphics[scale=0.57]{figures/standard}
  \vspace{0.25em}
  \caption{Sample charts built using derived abstractions; a scatter plot visualizing the Iris\\
    dataset with a title (left) and a pairplot comparing two Iris features (right).}
  \label{fig:standard}
\end{figure}

\section{Standard charts: Definining new abstractions}
\label{sec:abstractions}

The Compost library does not introduce a rich collection of standard charts and chart features
as most charting libraries. However, the functional domain-specific language design makes
it easy to define high-level chart features based on the low-level primitives of the core language.
To illustrate this, we give two examples.

First, one last remaining feature of the two charts in Figure~\ref{fig:charts} is a chart title.
This can be added to any chart using the following derived combinator:
%
\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{title}~t~s~=\kvd{overlay}~[\\
\quad \kvd{nest}_x~(\kvd{cont}~\num{0}), (\kvd{cont}~\num{100}),
  (\kvd{nest}_y~(\kvd{cont}~\num{0}), (\kvd{cont}~\num{15}), \\
\quad\quad \kvd{explicitScale}_x~(\kvd{continuous}~\num{0},\num{100}),~(\kvd{explicitScale}_y~(\kvd{continuous}~\num{0},\num{100}),\\
\quad\qquad \kvd{text}~\strf{\#000000}, (\kvd{cont}~\num{50}), (\kvd{cont}~\num{50}),~t)~)\\
\quad \kvd{nest}_x~(\kvd{cont}~\num{0}), (\kvd{cont}~\num{100}),
  (\kvd{nest}_y~(\kvd{cont}~\num{15}), (\kvd{cont}~\num{100}),~s)~]
\end{array}
\end{equation*}

\vspace{-0.5em}
\noindent
The \ident{title} combinator is a function defined using \fkvd{let} in the host language. It takes
a title $t$ and a shape $s$. It overlays two shapes. To position the title above the chart, the first
shape has an outer Y scale $\kvd{continuous}~\num{0},\num{15}$ while the second has an outer Y scale
$\kvd{continuous}~\num{15},\num{100}$. These are defined using the $\kvd{nest}_y$ primitive. Similarly,
the outer X scale of both is $\kvd{continuous}~\num{0},\num{100}$, defined using $\kvd{nest}_x$.

The second shape simply wraps the specified chart $s$ to which we are attaching the title. The first
positions the text title in the middle of the available space. To do so, we explicitly set the X and
Y scales inside the upper shape to continuous scales from $\num{0}$ to $\num{100}$ and then position
the text label in the middle, at a point $(\kvd{cont}~\num{50}), (\kvd{cont}~\num{50})$.
Figure~\ref{fig:standard} (left) shows a sample scatter plot chart with a title created using the
\ident{title} combinator.

A more complex type of chart that can be easily composed using the Compost primitives is the
pairplot chart from the seaborn Python library. % \cite
Pairplot visualizes pairwise relationships between features of a dataset.
An example using three features (sepal width, petal width, petal length) from the Iris dataset is shown
in Figure~\ref{fig:standard} (right). % \cite
A pairplot draws a grid of charts, each visualizing the relationship between two numerical features. For distinct
features, pairplot shows a scatter plot using one fature for X values and the other for Y values.
At the diagonal (when the fatures are the same), it draws a histogram of the values of the feature.
Additionally, a categorical feature can be used to determine the color of dots in the scatter plots.

To generate a pairplot, we use \kvd{nest} to overlay and align a grid of plots. Each of those
overlays a number of bubbles or filled shapes and adds left and bottom axis. As before, we use
\fkvd{let} to define a function and list comprehensions to generate individual chart elements.
We assume that \ident{data} is a list of rows, \ident{attrs} is a list of available attributes and
$\ident{get}~a~r$ obtains the attribute $a$ of a row $r$. We also assume the dataset contains the \str{color} attribute.
%
\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{pairplot}~\ident{attrs}~\ident{data}~=~\kvd{overlay}~[\\
\quad \fkvd{for}~x~\fkvd{in}~\ident{attrs}~\rightarrow~\fkvd{for}~y~\fkvd{in}~\ident{attrs}~\rightarrow\\
\qquad \kvd{nest}_x~(\kvd{cat}~x, 0), (\kvd{cat}~x, 1), (\kvd{nest}_y~(\kvd{cat}~y, 0), (\kvd{cat}~y, 1), \kvd{axis}_l~(\kvd{axis}_b\\
\quad\qquad (~\fkvd{if}~x \neq y~\fkvd{then}~\kvd{overlay}~[\;\fkvd{for}~r~\fkvd{in}~\ident{data}~\rightarrow~\\
\qquad\qquad \hspace{0.7em} \kvd{bubble}~(\ident{get}~\str{color}~r),~(\ident{get}~x~r), (\ident{get}~y~r), \num{1}, \num{1}~]\\
 \quad\qquad \hspace{0.7em} \fkvd{else}~\kvd{overlay}~[\;\fkvd{for}~x_1, x_2, y~\fkvd{in}~\ident{bins}~x~\ident{data}~\rightarrow~\\
\qquad\qquad \hspace{0.7em}  \kvd{fill}~\strf{\#808080}~[\; x_1, y, x_2, y, x_2, \num{0}, x_1, \num {0}\;]\;])))] \\
\end{array}
\end{equation*}

\vspace{-0.5em}
\noindent
As before, \kvd{nest} is essential for composing individual charts. Here, the points that determine
the locations of individual charts are categorical values defined by the attributes of the dataset.
The choice between two possible nested charts is made using the host language \fkvd{if} construct.
Scatter plots are generated by overlaying bubbles with X and Y coordinates obtained using $\ident{get}~x~r$
and $\ident{get}~y~r$. Histograms are composed from filled shapes. To obtain their locations, we use
a helper function $\ident{bins}~x~\ident{data}$, which returns a list of bins specified by a tripple
consisting of a lower and an upper range $x_1, x_2$ and the count $y$.

The example shows that the Compost is simple yet expressive. With just
a few lines of code, we are able to construct charts that, in other systems, require dedicated
libraries. The essential aspect of the language making this possible is the automatic inference
of scales and their mapping to the available space as well as the \kvd{nest} operation.
